---
title: "Example with simulated data"
author: "Julien Riou"
date: "04/12/2021"
output: html_document
---

```{r setup, include=FALSE}
# Setup
knitr::opts_chunk$set(echo = TRUE)
options(scipen=1, digits=1)
source("estimate_rr.R")
library(tidyverse)
set.seed(789)
theme_set(theme_bw())
```


## Simulate vaccination coverage

We start by simulating vaccination coverage for a year following a typical vaccination campaign.
We consider 5 age groups, and assume that the coverage will be higher in older age groups.
We also assume that the number of daily vaccinations starts high and decreases with time,
and scale the coverage so that it ends at 95% in the oldest age group.

```{r sim_cov, fig.width=5}
WEEKS = 1:52
AGE_GROUPS = 1:5
COV_MAX = .95
coveragedata = expand_grid(week=WEEKS, age_group=AGE_GROUPS) %>% 
  mutate(coverage=rpois(length(WEEKS)*length(AGE_GROUPS),10*1/week*age_group)) %>% 
  group_by(age_group) %>% 
  mutate(coverage=cumsum(coverage),
         coverage=age_group*coverage/max(coverage)/max(AGE_GROUPS)*COV_MAX)
```


```{r plot_cov, fig.width=5,fig.height=3,fig.align='center',fig.cap="Figure 1. Simulated vaccination coverage in time by age group."}
ggplot(coveragedata) +
  geom_line(aes(x=week,y=coverage,colour=factor(age_group))) +
  labs(colour="age_group") +
  scale_y_continuous(labels=scales::percent,limits=c(0,1))
```

## Simulate vaccination status among hospitalized

```{r sim_pars}
N = 5000
CHOSEN_RR = 10
```

We simulate typical surveillance data on hospitalized patients. We generate
`r N` patients and attribute to each a week of hospitalization and an age group uniformly. 
Using the coverage data, we link each patient with the vaccination coverage
at the corresponding time in the corresponding age group.

```{r sim_hosp}
survdata = tibble(id=1:N,
                  week=sample(WEEKS,N,replace=TRUE),
                  age_group=sample(AGE_GROUPS,N,replace=TRUE)) %>% 
  left_join(coveragedata,by=c("week","age_group"))
```

We now simulate the expected probability that each patient is vaccinated or not depending on 
(1) vaccination coverage and (2) a relative risk of hospitalization among non-vaccinated
compared to vaccinated of `r CHOSEN_RR`. For that we use the following equation:
$$
\Pr(V|H) = \frac{\Pr(V)}{\Pr(V) + RR\times(1-\Pr(V))}
$$
where $\Pr(V|H)$ is the probability of being vaccinated given hospitalization, $\Pr(V)$ is 
the probability of being vaccinated in the population (i.e. coverage), and $RR$ is 
relative risk of hospitalization among non-vaccinated compared to vaccinated ($RR=10$).
We then draw the actual vaccination status of each individual (`0` for not vaccinated, `1`
for vaccinated) from a Bernoulli distribution.

```{r sim_vacc}
survdata = survdata %>% 
  mutate(expected_vacc=coverage/(coverage+CHOSEN_RR*(1-coverage)),
         vacc=rbinom(N,size=1,prob=expected_vacc))
survdata
```

Aggregating over time, this results in an increasing trend in the proportion of vaccinated
among hospitalized cases.

```{r plot_vacc, fig.width=5,fig.height=3,fig.align='center',fig.cap="Figure 2. Simulated proportion of vaccinated among hospitalized patients."}
survdata %>% 
  dplyr::group_by(week) %>% 
  dplyr::summarise(prop_vaccinated=mean(vacc)) %>% 
  ggplot() +
  geom_line(aes(x=week,y=prop_vaccinated))+
  scale_y_continuous(labels=scales::percent)
```

## Estimate vaccine effectiveness

We can then apply the function `estimate_rr()` and directly estimate the $RR$
of hospitalization among non-vaccinated compared to vaccinated from individual
data on the vaccination status of hospitalized patients and the vaccine
coverage corresponding to each patient.

```{r est_rr}
EST_RR = estimate_rr(vacc=survdata$vacc, coverage=survdata$coverage)
```

We find a maximum likelihood estimate of `r EST_RR$rr` with a 
95% confidence interval of `r EST_RR$rr_lower` to  `r EST_RR$rr_upper`.
This can be compared to the value of `r CHOSEN_RR` used to simulate the data.

```{r plot_rr, fig.width=3,fig.height=3,fig.align='center',fig.cap="Figure 3. Estimated value of the relative risk of hospitalization among non-vaccinated compared to vaccinated compared to the value chosen to simulate the data."}
as_tibble(EST_RR) %>% 
  ggplot() +
  geom_pointrange(aes(x="Estimate",y=rr,ymin=rr_lower,ymax=rr_upper)) +
  geom_point(aes(x="Chosen value",y=CHOSEN_RR),colour="firebrick") +
  scale_y_continuous(limits=c(0,CHOSEN_RR*1.2)) +
  labs(x=NULL)
```


We can transform the $RR$ into a relative risk reduction with $1-1/RR$, resulting
in estimates of `r (1-1/EST_RR$rr)*100`% with a 95% confidence interval of `r (1-1/EST_RR$rr_lower)*100`% 
to  `r (1-1/EST_RR$rr_upper)*100`%. With the assumption that the differences between the
vaccinated and non-vaccinated groups (e.g. in terms of exposure or risk of hospitalization) 
only come from the vaccination status, this relative risk reduction can be interpreted
as vaccine effectiveness.

## Extensions

The function `estimate_rr()` allows the estimation of the overall $RR$ of hospitalization for non-vaccinated
compared to vaccinated. It can be applied to other outcomes, e.g. the $RR$ of death or the $RR$ of confirmed
case, provided that individual data with vaccination status is reliably collected.

If there are reasons to believe that the $RR$ varies according to individual characteristics (e.g. age or sex),
it is possible to estimate the variation in $RR$ according to covariates with function `estimate_rr_covariates()`.
Note that it is necessary to provide the covariates in matrix form. 

If vaccination status includes 3 possibilities instead of 2 (e.g. including several vaccine types or vaccination by time since
vaccination), then it is possible to estimate the $RR$ for different possibilities with function `estimate_rr_3groups()`.
It uses an extension of the equation of $\Pr(V|H)$ to obtain $\Pr(V_1|H)$, $\Pr(V_2|H)$ and $\Pr(V_3|H)$ and a categorical 
distribution instead of Bernoulli (see supplementary appendix). The 3 groups can be for instance not vaccinated, vaccinated
with vaccine 1 and vaccinated with vaccine 2. Or, restricting only to vaccinated people, the 3 groups can be vaccinated for
fewer than 3 months, 3-6 months and more than 6 months.